Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-04-08 paper/paper.tex
Reorder, and reduce the figures
Rework definitions, and reorder

\renewcommand{\floatpagefraction}{0.8}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}

@@

\begin{figure}[b] % first page if possible
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \begin{scope}[rotate=90]
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M8) at ($(330:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M9) at ($(270:1.5) + (M6)$) {\phantom{0}};

@@

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, in computer vision
\cite{DBLP:journals/jair/CookH94}, in the analysis of source code
\cite{DBLP:journals/tkde/DjokoCH97}, binary programs \cite{DBLP:conf/icics/GaoRS08}, and circuit
designs \cite{DBLP:journals/jair/CookH94}, in character recognition problems \cite{SIWEILU1991617},
and in many other domains \cite{Shasha:2002:AAT:543613.543620}, both directly and as a way of
measuring the similarity or difference between two graphs
\cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}. We illustrate two
variants of the problem in \cref{figure:mcsexample}.

Throughout this paper we will be working with graphs which may be directed, which may have labels
associated with vertices, edges, or both, and which may contain loops. To simplify notation, we will
work with a very broad definition: we define a \emph{graph} $G = (V, \ell, e)$ to be a set of
vertices $V$, a vertex labelling function $\ell : V \rightarrow \mathbb{N}$, and an edge labelling
function $e : V \times V \rightarrow \mathbb{N} \cup \{ \bot \}$. To represent graphs without vertex
(or edge) labels, we just pick our favourite number and give it to every vertex (or edge). We also
use a distinct label $\bot$ to represent the \emph{lack of} an edge between two vertices; vertices
$v$ and $w$ are said to be \emph{adjacent} if $e(v, w) \ne \bot$ or $e(w, v) \ne \bot$. We write
$\operatorname{V}(G)$ for $V$, and $\operatorname{N}(G, v)$ for the \emph{undirected neighbourhood}
of a vertex $v$ (that is, the set of vertices adjacent to $v$); a vertex is \emph{isolated} if its
neighbourhood is empty. A (weak) \emph{path} between two vertices $v$ and $w$ is a sequence of $n$
vertices $v = v_1, v_2, \ldots, v_n = w$ such that for each $m \in \{ 1 \ldots n - 1 \}$, $v_m$ is
adjacent to $v_{m + 1}$. A graph is (weakly) \emph{connected} if a path exists between each distinct
pair of vertices.

Given two graphs $G_1 = (V_1, \ell_1, e_1)$ and $G_2 = (V_2, \ell_2, e_2)$, we say two pairs of
vertices $(v_1, w_1) \in \operatorname{V}(G_1) \times \operatorname{V}(G_1)$ and $(v_2, w_2) \in
\operatorname{V}(G_2) \times \operatorname{V}(G_2)$ are \emph{compatible}, written $(v_1, w_1)
\simeq (v_2, w_2)$, if both $e_1(v_1, w_1) = e_2(v_2, w_2)$ and $e_1(w_1, v_1) = e_2(w_2, v_2)$. We
say two vertices $v \in \operatorname{V}(G_1)$ and $w \in \operatorname{V}(G_2)$ are
\emph{compatible}, written $v \simeq w$, if $\ell_1(v) = \ell_2(w)$ and $(v, v) \simeq (w, w)$.

An \emph{induced subgraph isomorphism} from a graph $G$ to a graph $H$ is an injective function $i :
\operatorname{V}(G) \rightarrow \operatorname{V}(H)$ such that for every vertex $v \in
\operatorname{V}(G)$, $v \simeq i(v)$, and for every pair of vertices $(v, w) \in
\operatorname{V}(G) \times \operatorname{V}(G)$, $(v, w) \simeq (i(v), i(w))$. A \emph{common
induced subgraph isomorphism} between two graphs $G_1$ and $G_2$ is a graph $G$, together with a
pair of induced subgraph isomorphisms $i_1 : G \rightarrow G_1$ and $i_2 : G \rightarrow G_2$. A
\emph{common induced subgraph} of $G_1$ and $G_2$ is a graph $G$ which has a common subgraph
isomorphism to $G_1$ and $G_2$. A \emph{maximum common (connected) induced subgraph} is a common
(connected) induced subgraph with the largest possible number of vertices.  (The distinction between
a common subgraph and a common subgraph isomorphism is relevant when counting distinct solutions,
which we do not consider in this paper; in \cref{figure:mcsexample} on the left, there is only one
maximum common subgraph, but four different maximum common subgraph isomorphisms are possible.)

There is also a non-induced version of these problems, which is commonly known as the \emph{maximum
common partial subgraph} problem: here, if $e(v, w) = \bot$, then $e_1(i_1(v), i_1(w))$ and
$e_2(i_2(v), i_2(w))$ may take any value, and we must maximise the number of non-$\bot$ vertex pairs
(i.e.\ edges) in $G$ instead (it is not entirely clear what the objective should be in directed
graphs, or when loops are present). We restrict ourselves to the induced version in this paper.

\subsection{Existing Approaches}

One approach to maximum common subgraph problems is via constraint programming
\cite{DBLP:conf/mco/VismaraV08,DBLP:conf/cp/NdiayeS11}. The models are based upon constructing an
injective mapping from the vertices of one graph (typically the smaller) to the other, in a similar
manner to subgraph isomorphism algorithms
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}. However, to
allow for a partial mapping, a special $\bot$ value is added to each domain (and so we are
constructing a maximum common subgraph isomorphism, where the common subgraph is implicitly the
subgraph of the first graph given by taking the vertices which are not assigned to $\bot$, and the
first subgraph isomorphism is the inclusion map). The objective is then to minimise the number of
variables given the $\bot$ value---for best results, a special propagator for the injectivity
condition is used, which simultaneously enforces that all variables taking non-$\bot$ are different,
and restricts the number of $\bot$ values which may be used based upon the objective
\cite{DBLP:conf/cp/PetitRB01}.  Unfortunately, the addition of $\bot$ also invalidates many of the
invariants used for subgraph isomorphism---we may no longer reason using the degree of vertices, for
example, and so maximum common subgraph problems can be much more challenging in practice.

Another widely studied approach ?? cite is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph \cite{LeviG}; we describe
this below. Conventional wisdom is that this model is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph model using a modern
algorithm.

(Other approaches have been tried, including dedicated algorithms ?? cite, mixed integer programming
\cite{DBLP:journals/anor/PivaS12}, and heuristics \cite{DBLP:journals/jcisd/EnglertK15}; SAT
encodings seem to struggle even for subgraph isomorphism \cite{UpcomingIJCAIPaper}.)

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05} to the model. Alternatively, we
could use a special branching rule to maintain connectedness during search ?? cite. Neither
technique seems directly viable with an association graph encoding---however, we show that it is
possible to adapt the combined branching and bounding rule used by modern clique algorithms to
preserve connectedness during search.

Our experiments suggest that each model and technique gives broadly similar performance, although
the association graph works better when graphs have edge labels (and we discuss why), and constraint
programming models are better when edges are unlabelled. For the constraint programming model, using
both a global constraint \emph{and} the branching rule in conjunction seems to be the safest
option---although it is not always fastest, it is very rarely substantially worse than just using
one or the other, and is often substantially better.

\section{Re-Evaluating the Clique Model}

@@

\subsection{Reduction to Maximum Clique in an Association Graph}

Explain association graph \cref{figure:association}, cite history \cite{LeviG}, connections to
microstructure \cite{DBLP:conf/aaai/Jegou93a}.

Discuss how edge labels, vertex labels, directed edges, etc, are handled. Note that we can generally
only handle very local side constraints, or injectivity (which is decomposable).

\begin{figure}[tb]
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \begin{scope}
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Ma) at (2, 10) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (Mb) at (0, 8) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (Mc) at (2, 8) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (Md) at (4, 8) {\phantom{0}};

            \node [right = 0 of Ma, font=\small] { \vphantom{0}a };
            \node [below = 0 of Mb, font=\small] { \vphantom{0}b };
            \node [below = 0 of Mc, font=\small] { \vphantom{0}c };
            \node [below = 0 of Md, font=\small] { \vphantom{0}d };

            \draw [very thick] (Mb) -- (Mc);
            \draw [very thick] (Mc) -- (Md);
            \draw [color=uofgsandstone!40] (Ma) -- (Mc);
            \draw [color=uofgsandstone!40] (Ma) to [out=150, in=80, looseness=8] (Ma);
            \draw [very thick] (Mb) to [out=150, in=80, looseness=8] (Mb);
        \end{scope}

        \begin{scope}
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M1) at (1, 2) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M2) at (3, 2) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (M3) at (1, 0) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M4) at (3, 0) {\phantom{0}};

            \node [left = 0 of M1, font=\small] { \vphantom{0}1 };
            \node [right = 0 of M2, font=\small] { \vphantom{0}2 };
            \node [left = 0 of M3, font=\small] { \vphantom{0}3 };
            \node [right = 0 of M4, font=\small] { \vphantom{0}4 };

            \draw [very thick] (M1) -- (M2);
            \draw [very thick] (M2) -- (M4);
            \draw [color=uofgsandstone!40] (M3) -- (M4);
            \draw [color=uofgsandstone!40] (M1) -- (M3);

            \draw [color=uofgsandstone!40] (M3) to [out=280, in=210, looseness=8] (M3);
            \draw [very thick] (M1) to [out=150, in=80, looseness=8] (M1);
        \end{scope}

        \begin{scope}[xshift=15cm]
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Ma1) at (2.6, 9) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Ma2) at (4.2, 9) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Ma3) at (5.8, 9) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Ma4) at (7.4, 9) {\phantom{0}};
            \node [above = 0 of Ma1, font=\small] { \vphantom{0}1 };
            \node [above = 0 of Ma2, font=\small] { \vphantom{0}2 };
            \node [above = 0 of Ma3, font=\small] { \vphantom{0}3 };
            \node [above = 0 of Ma4, font=\small] { \vphantom{0}4 };

            \draw [decorate, decoration={brace, raise=0.5cm}, very thick] (Ma1.north west) -- (Ma4.north east)
            node [midway, above=0.7cm] { $a \mapsto$ };

            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (Mb1) at (1, 2.6) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mb2) at (1, 4.2) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mb3) at (1, 5.8) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mb4) at (1, 7.4) {\phantom{0}};
            \node [left = 0 of Mb1, font=\small] { \vphantom{0}1 };
            \node [left = 0 of Mb2, font=\small] { \vphantom{0}2 };
            \node [left = 0 of Mb3, font=\small] { \vphantom{0}3 };
            \node [left = 0 of Mb4, font=\small] { \vphantom{0}4 };

            \draw [decorate, decoration={brace, raise=0.5cm}, very thick] (Mb1.south west) -- (Mb4.north west)
            node [midway, left=0.7cm] { $b \mapsto$ };

            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mc1) at (2.6, 1) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (Mc2) at (4.2, 1) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mc3) at (5.8, 1) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Mc4) at (7.4, 1) {\phantom{0}};
            \node [below = 0 of Mc1, font=\small] { \vphantom{0}1 };
            \node [below = 0 of Mc2, font=\small] { \vphantom{0}2 };
            \node [below = 0 of Mc3, font=\small] { \vphantom{0}3 };
            \node [below = 0 of Mc4, font=\small] { \vphantom{0}4 };

            \draw [decorate, decoration={brace, raise=0.5cm}, very thick] (Mc4.south east) -- (Mc1.south west)
            node [midway, below=0.7cm] { $c \mapsto$ };

            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Md1) at (9, 2.6) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Md2) at (9, 4.2) {\phantom{0}};
            \node[draw, circle, fill=white, draw=uofgsandstone!40, inner sep=0.5pt, font=\normalsize] (Md3) at (9, 5.8) {\phantom{0}};
            \node[draw, circle, fill=uofgcobalt, inner sep=0.5pt, font=\normalsize] (Md4) at (9, 7.4) {\phantom{0}};
            \node [right = 0 of Md1, font=\small] { \vphantom{0}1 };
            \node [right = 0 of Md2, font=\small] { \vphantom{0}2 };
            \node [right = 0 of Md3, font=\small] { \vphantom{0}3 };
            \node [right = 0 of Md4, font=\small] { \vphantom{0}4 };

            \draw [decorate, decoration={brace, raise=0.5cm}, very thick] (Md4.north east) -- (Md1.south east)
            node [midway, right=0.7cm] { $d \mapsto$ };

            \begin{scope}[on background layer]
                \draw [draw=uofgsandstone!40] ($(Ma1.south)!0.5!(Ma1)$) -- ($(Md4.west)!0.5!(Md4)$);
                \draw [draw=uofgsandstone!40] ($(Ma1.south)!0.5!(Ma1)$) -- ($(Mc2.north)!0.5!(Mc2)$);
                \draw [draw=uofgsandstone!40] ($(Ma3.south)!0.5!(Ma3)$) -- ($(Mc4.north)!0.5!(Mc4)$);
                \draw [draw=uofgsandstone!40] ($(Mb3.east)!0.5!(Mb3)$) -- ($(Mc4.north)!0.5!(Mc4)$);
                \draw [draw=uofgsandstone!40] ($(Md2.west)!0.5!(Md2)$) -- ($(Mc4.north)!0.5!(Mc4)$);
                \draw [draw=uofgsandstone!40] ($(Md2.west)!0.5!(Md2)$) -- ($(Ma3.south)!0.5!(Ma3)$);
                \draw [draw=uofgsandstone!40] ($(Md2.west)!0.5!(Md2)$) -- ($(Mb3.east)!0.5!(Mb3)$);
                \draw [very thick] ($(Mb1.east)!0.5!(Mb1)$) -- ($(Mc2.north)!0.5!(Mc2)$);
                \draw [very thick] ($(Mb1.east)!0.5!(Mb1)$) -- ($(Md4.west)!0.5!(Md4)$);
                \draw [very thick] ($(Mc2.north)!0.5!(Mc2)$) -- ($(Md4.west)!0.5!(Md4)$);
        \end{scope}
        \end{scope}

    \end{tikzpicture}

    \caption{A maximum common induced subgraph between the two graphs on the left has three
        vertices---one solution is highlighted. On the right, the association graph encoding: the
        highlighted clique of size three shows the same solution. In practice the isolated vertices
        (corresponding to assignments between incompatible vertices, which are unary constraints) would be
        omitted.}\label{figure:association}
\end{figure}

?? Discuss the partial version briefly, cite the reduction.

@@

\section{Find Maximum Common Connected Subgraphs}

>>>>

\renewcommand{\floatpagefraction}{0.8}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\newcommand{\christine}[1]{{\color{green} Christine:#1}}

@@

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, in computer vision
\cite{DBLP:journals/jair/CookH94}, in the analysis of source code
\cite{DBLP:journals/tkde/DjokoCH97}, binary programs \cite{DBLP:conf/icics/GaoRS08}, and circuit
designs \cite{DBLP:journals/jair/CookH94}, in character recognition problems \cite{SIWEILU1991617},
and in many other domains \cite{Shasha:2002:AAT:543613.543620}, both directly and as a way of
measuring the similarity or difference between two graphs
\cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}.

\paragraph{Definition of the Maximum Common (Connected) Subgraph Problem.}
We introduce definitions and algorithms on undirected and unlabelled graphs. The extension to directed or labelled graphs is straightforward and discussed in Section \ref{extension}. An undirected graph $G$ is defined by a finite set of nodes $N_G$ and a set of edges $E_G\subseteq N_G\times N_G$ such that each edge is an undirected pair of nodes.

Given a graph $G$, two nodes $n_s, n_e \in N_G$ are connected by a path in $G$ if there exists a sequence of nodes $<n_0,\ldots n_k>$ such that $n_0=n_s$, $n_k=n_e$ and each couple of successive nodes is connected by an edge, {\em i.e.}, $\forall i\in [1,k], \{n_{i-1},n_i\}\in E_G$. A graph $G$ is \emph{connected} if every pair of nodes is connected by a path.

A graph $G$ is \emph{isomorphic} to another graph $G'$ if there exists a bijective function $f:N_G\rightarrow N_{G'}$ which preserves edges, \emph{i.e.}, $\forall (u,v)\in N_G\times N_G, \{u,v\}\in E_G\Leftrightarrow \{f(u),f(v)\}\in E_{G'}$.

A {\em subgraph} is obtained by removing nodes, and all their incident edges, {\em i.e.}, $G'$ is a \emph{subgraph} of $G$ if $N_{G'}\subseteq N_G$ and $E_{G'}=E_G\cap(N_{G'}\times N_{G'})$.
In this definition, we only remove edges incident to removed nodes. This is usually called induced subgraph in the litterature. We may also allow to remove edges which are not incident to removed nodes, thus leading to partial subgraphs. In this paper, we only consider induced subgraphs. However, all algorithms may be extended to the partial case, as explained in \cite{DBLP:conf/cp/NdiayeS11}.

A \emph{common subgraph} of two graphs $G$ and $G'$ is a graph isomorphic to subgraphs of $G$ and $G'$. A \emph{common connected subgraph} is a common subgraph which is connected.
A \emph{Maximum Common Subgraph} (MCS) (resp. \emph{Maximum Common Connected Subgraph} (MCCS) is a common subgraph (resp. common connected subgraph) which has a maximum number of nodes. Examples of MCS and MCCS are given in Fig. \ref{figure:mcsexample}.

\begin{figure}[t] % first page if possible
    \centering
    \begin{tikzpicture}[scale=0.33]%{{{
        \begin{scope}[rotate=90]
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M1) at (90:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M2) at (150:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M3) at (30:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M4) at (210:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M5) at (330:1.5) {\phantom{0}};
            \node[draw, circle, draw=uofgsandstone!40, fill=white, inner sep=0.5pt, font=\normalsize] (M6) at (270:1.5) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M7) at ($(210:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M8) at ($(330:1.5) + (M6)$) {\phantom{0}};
            \node[draw, circle, fill=uofgsunshine, inner sep=0.5pt, font=\normalsize] (M9) at ($(270:1.5) + (M6)$) {\phantom{0}};

@@

\paragraph{Overview of the paper.} In Section \ref{existing}, we review existing approaches for solving the MCS problem, with a specific focus on Constraint Programming (CP)-based approaches and on clique-based approaches that reduce the problem to finding a maximum clique in an association graph.

Conventional wisdom is that clique-based approaches are less effective in practice, but previous experimental evaluations have used weak clique algorithms, or even maximal clique enumeration algorithms. 
Therefore, in Section \ref{eval1}, we  re-evaluate the clique-based approach using a modern algorithm, and we show that it outperforms CP on labelled graphs, and that it is competitive with CP on non labelled graphs.

Then, in Section \ref{mccs}, we consider the MCCS problem. For the CP approach, we may add a global connectedness constraint to the model. Alternatively, we may use a special branching rule to grow connected subgraphs only, as proposed in \cite{DBLP:conf/mco/VismaraV08}. These two techniques may be combined, and we experimentally show that the best results are obtained when combining them. 
When solving the MCCS problem with a clique-based approach, neither technique seems directly viable with an association graph encoding.
However, we show that it is possible to adapt the combined branching and bounding rule used by modern clique algorithms to preserve connectedness during search.
We compare the clique-based approach with the best CP variant for the MCCS, and we show that it outperforms CP on non labelled graphs, whereas it is outperformed by CP on labelled graphs.

\section{Existing Complete Approaches for MCS}\label{existing}

There exist two main approaches for solving MCS: The first approach (described in \ref{CP}) is based on CP; The second approach (described in \ref{clique}) is based on a reformulation of MCS into a maximum clique problem. Both approaches are described for non directed and non labelled graphs. Their extension to directed or labelled graphs is discussed in Section \ref{extension}

\subsection{CP model for MCS}
\label{CP}

McGregor \cite{McGreg82} proposed a Branch \& Bound approach: Each branch of the search tree corresponds to the matching of two nodes, and a  bounding function evaluates the number of nodes that still may be matched so that the current branch is pruned as soon as this bound becomes lower than the size of the largest known common subgraph. CP approaches may be viewed as enhancements of this Branch \& Bound approach.

Vismara and Valery  \cite{DBLP:conf/mco/VismaraV08} introduced a first CP model. Given two graphs $G$ and $G'$, this model associates a variable $x_u$ with every node $u$ of $G$, and the domain of this variable contains all nodes of $G'$, plus an additional value $\bot$: variable $x_u$ is assigned to $\bot$ if node $u$ is not matched to any node of $G'$; otherwise $x_u$ is assigned to the node of $G'$ it is matched with. Edge constraints are introduced in order to ensure that variable assignments preserve edges between matched nodes, {\em i.e.}, 
$\forall \{u,v\}\subseteq N_G, (x_u=\bot)\vee (x_v=\bot) \vee (\{u,v\}\in E_G\Leftrightarrow \{x_u,x_v\}\in E_{G'})$. Difference constraints are introduced in order to ensure that each node of $G'$ is assigned to at most one variable, {\em i.e.}, $\forall \{u,v\}\subseteq N_G, (x_u=\bot)\vee (x_v=\bot) \vee (x_u\neq x_v)$.

This CP model is improved in \cite{DBLP:conf/cp/NdiayeS11} by replacing binary difference constraints with a soft global allDifferent constraint which maximizes the number of $x_u$ variables that are assigned to values different from $\bot$, while ensuring they are all different when they are not assigned to $\bot$.
Different constraint propagation techniques are compared in \cite{DBLP:conf/cp/NdiayeS11}. The combination ``MAC+Bound'' (resp. ``FC+Bound'') obtains the best results on labelled (resp. non labelled) graphs and outperforms the approach of \cite{McGreg82} and the CP model of \cite{DBLP:conf/mco/VismaraV08}. The combination ``MAC+Bound'' maintains arc consistency~\cite{sabi94} of edge constraints, whereas the conbination ``FC+Bound'' simply performs forward-checking on these constraints. In both combinations, the ``Bound'' filtering checks whether it is possible to assign distinct values to enough $x_u$ variables to surpass the best cost found so far (it is a weaker version of GAC(\emph{softAllDiff}) \cite{peti01} which computes the maximum number of variables that can be assigned distinct values).

\subsection{Reformulation of MCS as a maximum clique problem}
\label{clique}

We may solve MCS by introducing an association graph and searching for a maximum clique in it \cite{bala86,dura99,DBLP:journals/jcamd/RaymondW02a}.

An association graph of two graphs $G$ and $G'$ is an undirected graph $A_{G,G'}$ whose set of nodes is $N_{A_{G,G'}} = N_G \times N_{G'}$. To avoid confusing nodes of $A_{G,G'}$ with initial graph nodes, nodes of $A_{G,G'}$ are called matching nodes, as each node $(u,u')$ of $A_{G,G'}$ denotes the matching of $u$ with $u'$. Edges of $A_{G,G'}$ connect matching nodes which are compatible, {\em i.e.}, $E_{A_{G,G'}} = \big\{\{(u,u'),(v,v')\}\subseteq N_{A_{G,G'}}$, $(u,u')$ and $(v,v')$ are compatible$\big\}$, where two matching nodes $(u,u')$ and $(v,v')$ of $N_{A_{G,G'}}$ are compatible if $u \neq v$ and $u' \neq v'$, and if they preserve edges (\emph{i.e.}, $\{u,v\} \in E_G \Leftrightarrow \{u',v'\} \in E_{G'}$).

A \emph{clique} is a subgraph whose nodes are all linked pairwise.
A clique is \emph{maximal} if it is not strictly included in any other clique, and it is \emph{maximum} if it is the largest clique of a given graph, with respect to the number of nodes.

A clique in $A_{G,G'}$ corresponds to a set of compatible matchings. Therefore, such a clique corresponds to a common subgraph, and a maximum clique of $A_{G,G'}$ is a MCS of $G$ and $G'$.
It follows that any method able to find a maximum clique in a graph can be used to solve the MCS problem.

Note that the association graph is a partial subgraph of the microstructure \cite{DBLP:conf/aaai/Jegou93a} associated with the CP model of  \cite{DBLP:conf/mco/VismaraV08}: The microstructure has more matching nodes than the association graph because it has a matching node $(u,\bot)$ for each node $u$ of $G$. Each clique of size $|N_G|$ in the microstructure corresponds to a common subgraph the size of which is defined by the number of nodes that do not contain $\bot$.

\subsection{Extension to Labeled or Directed Graphs}\label{extension}

In some applications, nodes and edges are associated with labels. We note $\lambda(u)$ and $\lambda(\{u,v\})$ the label of a node $u$ and an edge $\{u,v\}$, respectively. In this case, the common  subgraph must match components the labels of which are compatible, and we note $\simeq$ the compatibility relationship, {\em i.e.}, labels $\lambda_1$ and $\lambda_2$ are compatible if $\lambda_1\simeq\lambda_2$. This kind of label compatibility constraint is handled in a straightforward way in both CP and clique-based approaches:
\begin{itemize}
\item For CP, we restrict the domain of every variable $x_u$ to nodes with compatible labels, {\em i.e.}, $\forall u\in N_G, D(x_u) = \{\bot\}\cup \{u'\in N_{G'}, \lambda(u)\simeq\lambda(u')\}$. We also have to ensure that edge labels are preserved in edge constraints, {\em i.e.}, $\forall \{u,v\}\subseteq N_G, (x_u=\bot)\vee (x_v=\bot) \vee ((\{u,v\}\in E_G\Leftrightarrow \{x_u,x_v\}\in E_{G'}) \wedge (\lambda(\{u,v\}) \simeq \lambda(\{x_u,x_v\})))$.

\item For clique-based approaches, label compatibility is handled through the definition of the association graph, by restricting the set of matching nodes to pairs of nodes with compatible labels, {\em i.e.}, $N_{G_C} = \{(u,u')\in N_G\times N_{G'}, \lambda(u)\simeq\lambda(u')\}$. Also, edges of the association graph are restricted to pairs $\{(u,u'),(v,v')\}$ such that $\lambda(\{u,v\}) \simeq \lambda(\{u',v')\}$.
\end{itemize} 
Labels usually simplify the solution process, both for CP and clique-based approaches: Node labels reduce domain sizes for CP, and the number of matching nodes in the association graphs; Edge labels tighten edge constraints for CP, and make the association graph sparser for clique-based approaches.

The extension of MCS algorithms to directed graphs is straightforward: For CP, edge orientation is handled in edge constraints; for clique-based approaches, it is handled in the definition of the association graph edges. In both cases, adding orientations to edges usually simplifies the problem: It tightens edge constraints for CP, and it makes the association graph sparser for  clique-based approaches.

\section{Re-Evaluating the Clique Model}\label{eval1}

@@

@@

\section{Find Maximum Common Connected Subgraphs}\label{mccs}
